An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2020-07-15 manuscript/regional_befr_manuscript.tex
Updated refs.

Compared to forest ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire and herbivory, with African savannas possessing large herbivores absent from other savannas \citep{Sankaran2008, Levick2009}. Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance may reduce stem density, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. Reduced stem density reduces competition between individuals, possibly weakening the effect of diversity on ecosystem function via niche complementarity. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations under limiting environmental conditions such as low water availability \citep{}. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects. This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEF relationship.

@@

A small number of studies in southern African woodlands, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, with wide variation in precipitation, diurnal and annual temperature range, disturbance regime, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Group identification was conducted using unconstrained correspondence analysis, followed by clustering based on dominant ordination axes. Plot data used in this study occurred in four compositional vegetation types. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters. Cluster names were assigned post-hoc based on the dominant and indicator species in each cluster.

>>>>

Compared to forest ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire and herbivory, with African savannas possessing large herbivores absent from other savannas \citep{Sankaran2008, Levick2009}. Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{Dewees2010}. High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution. Disturbance may reduce stem density, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}. This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems. Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}. Reduced stem density reduces competition between individuals, possibly weakening the effect of diversity on ecosystem function via niche complementarity. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations under limiting environmental conditions such as low water availability. Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments. They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects. This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEF relationship.

@@

A small number of studies in southern African woodlands, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{Cardinale2009}. Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, with wide variation in precipitation, diurnal and annual temperature range, disturbance regime, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Group identification was conducted using unconstrained correspondence analysis, followed by hierarchical clustering based on dominant ordination axes. Plot data used in this study occurred in four compositional vegetation types. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters. Cluster names were assigned post-hoc based on the dominant and indicator species in each cluster.
